The Strucplot Framework---Visualizing Multi-way Contingency Tables

### 2006-01-04 trunk/vcd/inst/doc/strucplot.Rnw
several changes in structable framework
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@504 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
data(HairEyeColor)
@

@@

The strucplot framework in the {\sf R} package \pkg{vcd} visualizes
multi-way contingency tables and integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

@@

The strucplot framework is highly modularized: Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot. These are examples of `graphical appearance control'
(`grapcon') functions. They are created by generating functions, allowing
flexible parameterization and extensibility. 
The first part of their names (\code{\emph{group\_foo()}}) 
reflects the group they belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{\emph{struc\_foo()}}, 
\code{\emph{labeling\_foo()}}, and \code{\emph{legend\_foo()}}) 
directly produce graphical output (``add ink to the canvas''), whereas
the parameter functions
(created by \code{\emph{spacing\_foo()}} and \code{\emph{shading\_foo()}}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{\emph{struc\_foo()}} implement the core functionality, creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulas. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

\begin{figure}[p]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE>>=
<<Observed>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:observed}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE>>=
<<Expected>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}

@@

<<split1>>=
mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE))
@

@@

So far, we have visualized full tables. For objects of class \class{table},
conditioning on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is simply done by indexing.
However, subsetting \code{"structable"} objects is more restrictive 
because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
conditioning is only possible ``outside-in'':

<<subsetting,results=verbatim>>=
hec
hec["Male",]
hec[c("Male", "Brown"),]
hec["Male","Green"]
@

@@

pushViewport(viewport(layout.pos.col = 1))
mosaic(hec["Male"], margins = c(top = 2, 0), sub = "Male", newpage = FALSE)
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(hec["Female"], margins = c(top = 2, 0), sub = "Female", newpage = FALSE)
popViewport(2)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=5,width=12>>=
<<Variables>>
@
\caption{Distribution of hair and eye color, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\begin{figure}[p]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning>>
@
\caption{Mosaic plot for conditional independence structures.}
\label{fig:conditioning}
\end{center}
\end{figure}

@@

<<cotabplot>>=
cotabplot(~ Hair + Eye | Sex, data = hec, panel_args = list(margins = 3), labeling = labeling_left(clip = FALSE))
#cotabplot(~ Hair + Eye | Sex, data = hec, labeling_args = list(abbreviate = c(Eye = 2)))
@

@@

\noindent The \code{labeling\_args} argument ensures that `eye' levels
are abbreviated (see the vignette on labeling for detailed information).

>>>>

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
data(Titanic)
data(HairEyeColor)
data(PreSex)
data(UCBAdmissions)
art <- xtabs(~Treatment + Improved, data = Arthritis)
options(width=70)
@

@@

The strucplot framework in the {\sf R} package \pkg{vcd}, used for visualizing
multi-way contingency tables, integrates techniques such as
mosaic displays, association plots, and sieve plots. The main idea is to visualize
the tables' cells arranged in rectangular form. For multi-way tables,
the variables are nested into rows and columns using recursive
conditional splits, given the margins. The result is a 
`flat' representation that can be visualized in
ways similar to a two-dimensional table.
This principle defines a class of conditional displays which allows
for granular control of graphical appearance aspects, including:

@@

The strucplot framework is highly modularized: Figure \ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot. These are examples of `graphical appearance control'
(`grapcon') functions. They are created by generating functions
(`grapcon generators'), allowing
flexible parameterization and extensibility (Figure \ref{fig:struc}
only shows the generators).
The first part of the generator names (\code{\emph{group\_foo}()})
reflects the group they belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\emph{foo}()}, 
\code{labeling\_\emph{foo}()}, and \code{legend\_\emph{foo}()}) 
directly produce graphical output (``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\emph{foo}()} and \code{shading\_\emph{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\emph{foo}()} implement the core functionality, creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure \ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulas. Finally, on the fourth
level, there are `related' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

\setkeys{Gin}{width=0.75\textwidth}
\begin{figure}[p]
\begin{center}
<<Observedfig,fig=TRUE,echo=FALSE>>=
<<Observed>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data.}
\label{fig:observed}
\end{center}
\end{figure}

@@

\begin{figure}[p]
\begin{center}
<<Expectedfig,fig=TRUE,echo=FALSE>>=
<<Expected>>
@
\caption{Mosaic plot for the \data{HairEyeColor} data (expected values).}
\label{fig:expected}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

<<echo=FALSE>>=
options(width=60)
@

<<split1>>=
mosaic(hec, split_vertical = c(TRUE, FALSE, TRUE), 
labeling_args = list(abbreviate = c(Eye = 3)))
@

<<echo=FALSE>>=
options(width=70)
@

@@

So far, we have visualized full tables. Subtables can be selected in a
similar way than for objects of class \class{table} using
indexing. Note, however, that subsetting of \code{"structable"} objects 
is more restrictive because of their inherent conditional structure. Since the variables
on both the row and the columns side are nested, 
subsetting is only possible ``outside-in'', that is, indexing operates
on blocks defined by the variable levels:

<<subsetting,results=verbatim>>=
hec
hec["Male",]
hec["Male", c("Brown","Green")]
@

\noindent \emph{Conditioning} on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is done using the \code{[[} operator. %]]
Here again, the sequence of conditioning levels is restricted by the
hierarchical structure of the \class{"structable"} object:

<<conditioning,results=verbatim>>=
hec[["Male",]]
hec[[c("Male", "Red"),]]
hec[["Male","Green"]]
@

@@

pushViewport(viewport(layout.pos.col = 1))
mosaic(hec[["Male"]], margins = c(left = 2.5, top = 2.5, 0), sub = "Male", newpage = FALSE)
popViewport()

pushViewport(viewport(layout.pos.col = 2))
mosaic(hec[["Female"]], margins = c(top = 2.5, 0), sub = "Female", newpage = FALSE)
popViewport(2)
@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Variablesfig,fig=TRUE,echo=FALSE,height=6,width=12>>=
<<Variables>>
@
\caption{Distribution of hair and eye color, given gender.}
\label{fig:parttable}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<Conditioningfig,fig=TRUE,echo=FALSE,width=12>>=
<<Conditioning>>
@
\caption{Mosaic plot for conditional independence structures.}
\label{fig:conditioning}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7}

@@

<<cotabplot>>=
cotabplot(~ Hair + Eye | Sex, data = hec, panel_args = list(margins = 3), labeling = labeling_left(clip = FALSE))
@

@@

\noindent The \code{labeling\_args} argument modifies the labels'
appearance: here, to be left-aligned and unclipped
(see the separate vignette: ``Labeling in the Strucplot Framework'' 
for detailed information).
